Quotients of the magmatic operad:\\ lattice structures and convergent rewrite systems

### 2018-09-05 Introduction.tex
MAJ

To explore this vast research area, we select to pursue in this paper
the following directions. First, we consider the very general set of the
quotients of $\Mag$ seen as an operad in the category of vector spaces.
We show that these operads form a lattice, wherein its partial order
relation is defined from the existence of operad morphisms
(Theorem~\ref{thm:lattice_structure_of_QMag}). We also provide a
Grassmann formula analog relating the Hilbert series of the operads of
the lattice together with their infima and suprema
(Theorem~\ref{thm:Grassmann_formula_for_Hilbert_series_of_QMag}).
Besides, we focus on a special kind of quotients of $\Mag$, denoted by
$\CAs{\gamma}$, defined by equating the left and right comb binary trees
of a fixed degree $\gamma \geq 1$. These operads are called comb
associative operads. For instance, $\CAs{3}$ is the operad describing
the category of the algebras equipped with a binary product $\Product$
subjected to the relation
\begin{equation}
    \left(\left(x_1 \Product x_2\right) \Product x_3\right) \Product x_4
    =
    x_1 \Product \left(x_2 \Product \left(x_3 \Product x_4\right)\right).
\end{equation}
This is a kind of associativity law in higher degree $\gamma = 3$ so
that $\CAs{3}$ can be seen as a generalization of $\As$ (which is equal
to $\CAs{2}$). We first provide general results about the operads
$\CAs{\gamma}$. In particular, we show that these operads form a lattice
contained in the lattice of quotients of $\Mag$ aforementioned
(Theorems~\ref{thm:lattice_CAs} and~\ref{thm:inclusion_lattice_CAs}). We
focus in particular on the study of $\CAs{3}$. Observe that the
congruence of this operad is generated by an equivalence class of trees
which is not an interval for the Tamari order. As preliminary computer
experiments show, $\CAs{3}$ has oscillating first dimensions
(see~\eqref{equ:dimensions_CAs_3}), what is rather unusual among all
known operads. We provide a convergent orientation of the space of
relations of $\CAs{3}$ (Theorem~\ref{thm:convergent_rewrite_rule_CAs_3}),
a description of a basis of the operad, and prove that its Hilbert series
is rational.
%
\Todo{S: et parler de la réalisation si Christophe l'ajoute ?}
%
For all these, we use rewrite systems on trees~\cite{BN98} and the
Buchberger algorithm for operads~\cite{DK10}. We expose some
experimental results obtained with the help of the computer for some
operads $\CAs{\gamma}$ with $\gamma \geq 4$. Finally, we continue the
investigation of the quotients of $\Mag$ by regarding the quotients of
$\Mag$ obtained by equating two trees of degree $3$. This leads to ten
quotient operads of $\Mag$. We provide for some of these combinatorial
realizations, mostly in terms of integer compositions.
\medbreak

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Acknoledgements}
The authors wish to thank Maxime Lucas for helpfull discussions and
Vladimir Dotsenko for his mark of interest to this work.
\medbreak

>>>>

To explore this vast research area, we select to pursue in this paper
the following directions. First, we consider the very general set of the
quotients of $\Mag$ seen as an operad in the category of vector spaces.
We show that these operads form a lattice, wherein its partial order
relation is defined from the existence of operad morphisms
(Theorem~\ref{thm:lattice_structure_of_QMag}). We also provide a
Grassmann formula analog relating the Hilbert series of the operads of
the lattice together with their lower-bound and upper-bound
(Theorem~\ref{thm:Grassmann_formula_for_Hilbert_series_of_QMag}).
Besides, we focus on a special kind of quotients of $\Mag$, denoted by
$\CAs{\gamma}$, defined by equating the left and right comb binary trees
of a fixed degree $\gamma \geq 1$.
Observe that since $\CAs{2}=\As$, the operads $\CAs{\gamma}$ can be seen
as generalizations of $\As$. These operads are called comb
associative operads. For instance, $\CAs{3}$ is the operad describing
the category of the algebras equipped with a binary product $\Product$
subjected to the relation
\begin{equation}
    \left(\left(x_1 \Product x_2\right) \Product x_3\right) \Product x_4
    =
    x_1 \Product \left(x_2 \Product \left(x_3 \Product x_4\right)\right).
\end{equation}
We first provide general results about the
operads $\CAs{\gamma}$. In particular, we show that these operads form a
a lattice which embeds as a poset in the lattice of the quotients of
$\Mag$ aforementioned
(Theorems~\ref{thm:lattice_CAs} and~\ref{thm:inclusion_lattice_CAs}). We
focus in particular on the study of $\CAs{3}$. Observe that the
congruence of this operad is generated by an equivalence class of trees
which is not an interval for the Tamari order. As preliminary computer
experiments show, $\CAs{3}$ has oscillating first dimensions
(see~\eqref{equ:dimensions_CAs_3}), what is rather unusual among all
known operads. We provide a convergent orientation of the space of
relations of $\CAs{3}$ (Theorem~\ref{thm:convergent_rewrite_rule_CAs_3}),
a description of a basis of the operad, and prove that its Hilbert series
is rational.
%
\Todo{Parler des formes normales}
%
For all these, we use rewrite systems on trees~\cite{BN98} and the
Buchberger algorithm for operads~\cite{DK10}. We expose some
experimental results obtained with the help of the computer for some
operads $\CAs{\gamma}$ with $\gamma \geq 4$. Finally, we continue the
investigation of the quotients of $\Mag$ by regarding the quotients of
$\Mag$ obtained by equating two trees of degree $3$. This leads to ten
quotient operads of $\Mag$. We provide for some of these combinatorial
realizations, mostly in terms of integer compositions.
\medbreak

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Acknoledgements}
The authors wish to thank Maxime Lucas for helpfull discussions and
Vladimir Dotsenko for his marks of interest and his bibliographic
suggestions.
\medbreak

### 2018-09-05 Operads_rewrite.tex
MAJ

Let us now go back on operads. Let $\Congr$ be an operad congruence of
$\Mag$. If $\Tfr$ is a binary tree, we denote by $[\Tfr]_{\Congr}$ the
$\Congr$-equivalence class of $\Tfr$. By definition, $[\Tfr]_{\Congr}$
is an element of the quotient operad
\begin{equation} \label{equ:quotient_operad_mag}
    \Oca := \Mag/_{\Congr}.
\end{equation}
A set of rewrite rules $\Rew$ on $\Mag$ is an \Def{orientation} of
$\Congr$ if $\RewContextRST$ and $\Congr$ are equal as binary relations,
where $\RewContext$ is the rewrite relation induced by $\Rew$. Moreover,
$\Rew$ is a \Def{convergent} (resp. \Def{terminating}, \Def{confluent})
orientation of $\Congr$ if $\RewContext$ is convergent (resp.
terminating, confluent). A \Def{presentation} of the quotient operad
$\Oca$ of the form~\eqref{equ:quotient_operad_mag} is the data of a
generating set for the operad congruence $\Congr$. Observe that
any orientation of $\Congr$ is a presentation of $\Oca$, so that the
above nomenclature (\Def{convergent}, \Def{terminating}, and
\Def{confluent}) still holds for presentations. A presentation is said
to be \Def{finite} if it is a finite set.
\medbreak

@@

\Todo{S: je crois qu'il faudrait préciser les choix que l'on fait aux
étapes (3), (4) et (5).}

>>>>

Let us now go back on operads. Let $\Congr$ be an operad congruence of
$\Mag$. If $\Tfr$ is a binary tree, we denote by $[\Tfr]_{\Congr}$ the
$\Congr$-equivalence class of $\Tfr$. By definition, $[\Tfr]_{\Congr}$
is an element of the quotient operad
\begin{equation} \label{equ:quotient_operad_mag}
    \Oca := \Mag/_{\Congr}.
\end{equation}
A set of rewrite rules $\Rew$ on $\Mag$ is an \Def{orientation} of
$\Congr$ if $\RewContextRST$ and $\Congr$ are equal as binary relations,
where $\RewContext$ is the rewrite relation induced by $\Rew$. Moreover,
$\Rew$ is a \Def{convergent} (resp. \Def{terminating}, \Def{confluent})
orientation of $\Congr$ if $\RewContext$ is convergent (resp.
terminating, confluent). We call \Def{presentation} of a quotient
operad $\Oca$ of the form~\eqref{equ:quotient_operad_mag} the data of
a generating set for the operad congruence $\Congr$. Observe that
any orientation of $\Congr$ is a presentation of $\Oca$, so that the
above nomenclature (\Def{convergent}, \Def{terminating}, and
\Def{confluent}) still holds for presentations. A presentation is said
to be \Def{finite} if it is a finite set.
\medbreak

@@
